Stability and Change in Personal Culture

### 2020-02-14 manuscript9.Rmd
final submission revisions

Despite their diversity, models of aggregate attitude and behavior change often implicitly invoke one of two broad models of individual change with strong connections to theories in cultural sociology. The first of these models is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions on attitude formation. This model is rooted in a broadly pragmatist approach to action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas 1996, Gross 2009, Swidler 2001). The second is a _settled dispositions_ model, which emphasizes the continuing influence of durable dispositions acquired early in life. This model has affinities with the Bourdieusian tradition, which de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016). These two models represent different approaches to understanding how people come to hold diverse forms of _personal culture_, or the declarative and non-delcarative attitudes, worldviews, values, dispositions, and associations that manifest at the individual level [@lizardo17].

@@

```{r dag, message=FALSE, fig.width=6.8, fig.cap="Graphical representation of active updating and settled dispositions models over time."}
library(ggdag)
library(dagitty)
aum <- 
  dagitty("dag {
        y1 -> y2 -> y3 ;
        v1 -> y1 ;
        v2 -> y2 ;
        v3 -> y3 
        }")
coordinates( aum ) <-
  list( x=c(y1 = 1, y2 = 2, y3 = 3, v1 = 1, v2 = 2, v3 = 3),
        y=c(y1 = 2, y2 = 2, y3 = 2, v1 = 1, v2 = 1, v3 = 1) )

gridExtra::grid.arrange(aum_dag, sdm_dag, ncol = 2)

@@

Alongside ideological identificaton, moral issues and general political sentiments might serve as core organizing principles, but what these beliefs are will not be consistent across people [@baldassarigelman]. For example, some people might hold strong, consistent views on abortion while others care about the amount of government intervention in the economy. Because of this heterogeneity, these beliefs should have a moderate level of consistency, since those people who do not value them should report them inconsistently. There should be no evidence of updating in these beliefs, as previous work suggests that people do not bring these attitudes in line with other beliefs, even if they are discordant.

@@

Our analysis proceeds in three steps to answer our three research questions. First, we evaluate the overall evidence in support of the active updating model. To do this, we compare for each item the Bayesian Information Criteria (BIC) of a model estimated using Equation \@ref(eq:MODEL) with a free estimate of $\phi$ to a model that constrains $\phi= .5$. We calculate the posterior probability that the model with the free parameter fits the data better. If the model with the constraint is preferred, then we conclude that both wave 1 and wave 2 are equally good predictors of wave 3, meaning there is no evidence that respondents are actively updating on that item.

@@

@@

@@

@@

@@

@@

While people tend to be inconsistent in their responses to spending priorities, views on other policy questions with both high and low salience are reported quite consistently. Views on abortion, especially when considered as a scale, are consistently reported. As with all other scales, aggregating the composite items increases consistency but does not affect the estimate of the amount of active updating, validating our method's ability to separate persisting and non-persisting change. As discussed previously, views on gay marriage display high levels of active updating. In contrast, the item asking about support for legalizing recreational marijuana use is reported with consistency but weak active updating. This is notable, as public opinion and policies regarding both have shifted considerably in recent years. Previous work suggests that change in both could be driven by the same underlying process [@schnabelsevell]. Our findings suggest that, at least in recent years, many individuals have changed their views on gay marriage, but changing views on marijuana have been driven primarily by cohort replacement.

@@

Figures \ref{fig:phi_relig} through \ref{fig:phi_racgen} plot $\phi$ estimates for all items included in the analysis, grouped by subject material.

@@

@@

$\phi$ estimates for coarsened items are presented in Figures \ref{fig:coarse_relig} through \ref{fig:coarse_racgen}. If scales had symmetrical scales with no clear midpoint (e.g., strongly agree/agree/disagree/strongly disagree), we coarsened those scales to two points (agree/disagree). If scales included a clear midpoint, we included that and coarsened responses on either side. For items on large scales such as hours of TV watched, we split responses into greater than the median or less than or equal to the median. If changes between ends of each scale are more persistent than changes within ends of these scales, then $\phi$ estimates of coarsened item should be higher than estimates for items with more response options.

>>>>

Despite their diversity, models of aggregate attitude and behavior change often implicitly invoke one of two broad models of individual change with strong connections to theories in cultural sociology. The first of these models is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions on attitude formation. This model is rooted in a broadly pragmatist approach to action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas 1996, Gross 2009, Swidler 2001). The second is a _settled dispositions_ model, which emphasizes the continuing influence of durable dispositions acquired early in life. This model has affinities with the Bourdieusian tradition, which de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016). These two models represent different approaches to understanding how people come to hold diverse forms of _personal culture_, or the declarative and non-declarative attitudes, worldviews, values, dispositions, and associations that manifest at the individual level [@lizardo17].

@@

[Figure \@ref(fig:dag) about here]

```{r}
library(ggdag)
library(dagitty)
```

@@

Alongside ideological identification, moral issues and general political sentiments might serve as core organizing principles, but what these beliefs are will not be consistent across people [@baldassarigelman]. For example, some people might hold strong, consistent views on abortion while others care about the amount of government intervention in the economy. Because of this heterogeneity, these beliefs should have a moderate level of consistency, since those people who do not value them should report them inconsistently. There should be no evidence of updating in these beliefs, as previous work suggests that people do not bring these attitudes in line with other beliefs, even if they are discordant.

@@

Our analysis proceeds in three steps to answer our three research questions. First, we evaluate the overall evidence in support of the active updating model. To do this, we compare for each item the Bayesian Information Criteria (BIC) of a model estimated using Equation \@ref(eq:MODEL) with a free estimate of $\phi$ to a model that constrains $\phi= .5$. We calculate the posterior probability that the model with the free parameter fits the data better. If the model with the constraint is preferred, then we conclude that both wave 1 and wave 2 are equally good predictors of wave 3, meaning there is no evidence that respondents are actively updating on that item.[^link]

[^link]: All data used in this analysis is publicly available through the General Social Survey at the National Opinion Research Center at http://gss.norc.org/. R code for how we cleaned the GSS panels, all analyses performed in this paper, and all figures created in this paper can be found at http://github.com/krkiley/panel_change.

@@

[Figure \@ref(fig:phihist) about here.]

@@

[Figure \@ref(fig:agegroupcomparison) about here.]

@@

[Figure \@ref(fig:phibetascatter) about here.]

@@

[Figure \@ref(fig:summary) about here.]

@@

[Figure \@ref(fig:polvariables) about here.]

@@

While people tend to be inconsistent in their responses to spending priorities, views on other policy questions with both high and low salience are reported quite consistently. Views on abortion, especially when considered as a scale, are consistently reported. As with all other scales, aggregating the composite items increases consistency but does not affect the estimate of the amount of active updating, validating our method's ability to separate persisting and non-persisting change. As discussed previously, views on gay marriage display high levels of active updating. In contrast, the item asking about support for legalizing marijuana use is reported with consistency but weak active updating. This is notable, as public opinion and policies regarding both have shifted considerably in recent years. Previous work suggests that change in both could be driven by the same underlying process [@schnabelsevell]. Our findings suggest that, at least in recent years, many individuals have changed their views on gay marriage, but changing views on marijuana have been driven primarily by cohort replacement.

@@

Figures \@ref(fig:religphi) through \@ref(fig:racgenphi) plot $\phi$ estimates for all items included in the analysis, grouped by subject material.

[Figures \@ref(fig:religphi) through \@ref(fig:racgenphi) about here.]

@@

Estiates of $\phi$ for coarsened items are presented in Figures \@ref(fig:coarserelig) through \@ref(fig:coarseracgen). If scales had symmetrical scales with no clear midpoint (e.g., strongly agree/agree/disagree/strongly disagree), we coarsened those scales to two points (agree/disagree). If scales included a clear midpoint, we included that and coarsened responses on either side. For items on large scales such as hours of TV watched, we split responses into greater than the median or less than or equal to the median. If changes between ends of each scale are more persistent than changes within ends of these scales, then $\phi$ estimates of coarsened item should be higher than estimates for items with more response options.

[Figures \@ref(fig:coarserelig) through \@ref(fig:coarseracgen) about here.]

```{r dag, message=FALSE, fig.width=6.8, fig.cap="Graphical representation of active updating and settled dispositions models over time."}
aum <- 
  dagitty("dag {
        y1 -> y2 -> y3 ;
        v1 -> y1 ;
        v2 -> y2 ;
        v3 -> y3 
        }")
coordinates( aum ) <-
  list( x=c(y1 = 1, y2 = 2, y3 = 3, v1 = 1, v2 = 2, v3 = 3),
        y=c(y1 = 2, y2 = 2, y3 = 2, v1 = 1, v2 = 1, v3 = 1) )

gridExtra::grid.arrange(aum_dag, sdm_dag, ncol = 2)
```

@@
